Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-06 boettiger.tex
ironing out text

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,preprint,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}

\journal{\tiny Journal of the Royal Society Interface} 
\begin{document}
\begin{frontmatter}
  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
  \author[cpb]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  \author[esp]{Alan Hastings}
  %\author[info]{}
  %\author[davis]{}
  \cortext[cor1]{Corresponding author.}
  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}

@@

@@

\cb{Need more introduction defining critical transitions, early warning signals etc?}

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption %often unstated?
that the system dynamics contain a saddle-node bifurcation and
looks for patterns that would be associated with approaching this bifurcation point.
The typical pattern has been a trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
% consider additional citations
While attractive for their simplicity, such approaches are subject to numerous challenges.
In this paper we argue for a model-based approach, 
and describe how this can be done in a way that best addresses these difficulties.

\subsubsection*{Arbitrary windows}
While superficially simple to use, the calculation of these statistics is subject to 
several arbitrary decisions and additional assumptions,
such as the choice of window size over which the statistic is computed,
the choice of detrending,
and whether and by how much the consecutive time-windows used in the calculation should overlap.  
\citet{Lenton2012} compare the performance of different approaches to these issues,
showing that choice of window and method used for detrending indeed influence the results,
and each approach considered has pros and cons.  
Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
it is difficult to evaluate and compare between such methods.  
A model-based approach side-steps these issue entirely.  
\cb{
 Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
 and that it assumes stationarity, which is precisely the thing we're looking to change?  
 Similarly could mention that the detrending approaches can remove the signal?
 We could also mention that some popular approaches such as autocorrelation require evenly-spaced points,
 and interpolating is incorrect, introducing fictitious correlations?
}

\subsubsection*{No quantitative measures}
The warning signal, ``an increase in statistic $x$,'' is not quantitative,
making it difficult to compare between signals or to
attribute a statistical significance to the detection.
Some authors have suggested Kendall's correlation coefficient to quantify an increase~\citep{Dakos2008, Dakos20011} 
in autocorrelation or variance.  
Other measures of increase, such as Pearson's, have also been proposed~\citep{Drake2010}.  
Much of the literature using summary statistics simply forgoes quantifying
the increase or estimating significance.  
While this may be satisfactory in experimental systems with controls and replicates to compare to,~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
any real-world application of early warning methods will not have the luxury of controls and replicates.
In such cases a quantitative definition of detection is essential.  
\cb{ We will see that correlation tests in summary statistics are an ineffective way to do this.  }

\subsubsection*{Problematic null models}
Specifying an appropriate null model is also difficult.  
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.  
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent, but this is
an assumption that is violated by the very experimental design:
since one variable is time and the other a moving time-averaged window.  
Under such a test any adequately large data set will find a significant result, regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
this destroys the natural autocorrelation in the time series.  
\cb{I feel citing \citet{Dakos2011} and \citet{Dakos2008} for these two examples would be impolite\dots} 
Parametric null models have been more promising, and have brought us closer to a model-based approach where assumptions are explicit.  
For instance, autoregressive models~\citet{Dakos2008} have been proposed, which are first parameterized from the data.  
\cb{is this next line okay to leave in?}
Estimating the null models directly from the data using a maximum likelihood framework, will provide the most powerful basis for this test.  
\citet{Seekell2011} proposes a conditional heteroscedaticity as a summary statistic particularly because probability models for the null distribution are readily available.\cb{Though as a measure of changing variance, it suffers from the weaknesses mentioned above, but no need to mention that.}

\cb{This is a particular concern given the tendency for warning signals to be found only in the largest data sets, 
which has been taken as evidence that large data sets are required for successful detection,
not that large data sets have higher false-positive rates under certain metrics.}
\cb{A similar problem exists with the proposal that large deviations from the running variance,
i.e. in excess of 2 standard deviations, be taken as evidence of an increase.
This likewise guarantees the false positive rate goes to unity as the number of observations increases.}

\subsubsection*{Hidden assumptions}
Assumes that the system sits effectively at equilibrium around a stable state  
and tests whether the system may be experiencing the slow change of a parameter that is weakening the stability of that state,
moving the system towards a saddle-node bifurcation.  
As such processes may be common in natural systems this is an important scenario to consider,
but without the clarity of assumptions made in a modeling approach there is a risk that
early warning signs will be used in situations where they do not apply.

%Differing system dynamics
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in summary-statistic based approach.  
For instance, variance may increase for reasons that 
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.   
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by 
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

\cb{ do we need the energy diagram figure to emphasize this is our assumption?}
%Differing causes of the transition.
Even when we can exclude or ignore such dynamics and 
restrict ourselves to systems that can support a saddle-node bifurcation,
any early warning signals approach must further assume that a changing parameter 
has brought the system closer to the bifurcation.  This excludes at least three alternative
explanations for the transition.  
The first is that a large perturbation of the system state 
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.  
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast. 
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}  
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.  
The third scenario is that the system does pass through a saddle-node bifurcation, 
but rather than gradually and monotonically approaching the critical point, the 
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection{Summary-statistic approaches have less statistical power.}
Methods for the detection of early warning signals are continually challenged by 
inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
The challenge of extrapolating into the future across a dramatic shift is immense;
there is essentially no historical president for successfully \& reliably predicting regime shifts
outside of controlled settings. % \cite Tetlock?
The signals one attempts to detect with these approaches are extremely weak.
Despite the widespread recognition of the this need for large data sets, 
there has been very few studies quantitative studies of power looking at how much data is actually required~\citep{Contamin2009}.
\cb{this is the only example that claims to measure statistical power.  I find it incredible that Ellison, contrary to his own ecological statistics textbook, decides to just invent an arbitrary statistic (log ratio of the maximum value in the statistic seen before and after onset of the perturbation) and call this a measure of ``statistical power.''}
The Neyman-Pearson Lemma guarantees us that the most powerful test between hypotheses compares the 
likelihood that the data was produced under each. This requires a model-based approach, which,
as we will see, can have substantially more power to detect early warning signals than summary-statistic based approaches.

@@

\subsection{Early warning signals as model choice}
It may be useful to think of the detection of early warning signals as a problem of model choice, rather than one of pattern recognition.  
The model-choice based approach attempts to frame each of the possible scenarios as structurally different equations,
making the assumptions explicit.  
In any model choice problem, it is important to identify the goal of the exercise -- such as the ability to generalize, to immitate reality, or to predict~\citep{Levins1966}.  
In this case generality is more important than realism or predictive capability: 
we will write down a general model that is capable of approximating the wide class of models that experience a critical transition through a saddle node bifurcation,
and another general model that is capable of representing these same systems when they are not approaching a bifurcation.  
These may be thought of as the hypothesis and null hypothesis,
though they are in fact compound hypotheses: in each case, we must first estimate the model parameters from the data before we can compare the models.  
In this approach it is not assumed that ``reality'' is included in the models being tested per se, 
but that one of the models is a better approximation of the true dynamics than the other.  
System whose dynamics violate the assumptions common to both models, 
such as in the examples of ~\citet{Hastings2009} where systems exhibit sudden transitions without warning, 
fall outside the set of cases where this approach would be valid; 
though the inability of either model to match the system dynamics could be an indication of such a violation.

\subsection{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter, which may be slowly varying in time.  
Transforming this canonical form to allow for an arbitrary mean $\theta$,
the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi(r_t)$.
We expand around the fixed point and express in terms of the stochastic dynamics:

@@

\subsection{Likelihood calculations}\label{likelihood}
The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
\begin{equation}
\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
\end{equation}
For~\eqref{LSN} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
\begin{align}
 \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
  \label{general_moments}
\end{align}

@@

\subsection{Comparing Models}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

Let $L_0$ be the likelihood function for model 0, 
let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) 
be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; 
and define $L_1$, $\theta_1$, $L_1$ similarly for model 1. 
The statistic we will use is $\delta$, 
defined to be twice the difference in log likelihood of observing the data under the two MLE models,
$\delta = -2 (\log L_0 - \log L_1 )$.  
This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.

\section{On a hypothesis-testing framework}\label{Dakos}
\cb{Here's a longer section on hypothesis testing that was originally in the appendix.  One of the Ecology Letters reviewers seemed very upset that this information was in the appendix and not in the main text, so here it is.  This section is quite specific in references to ~\citet{Dakos2008}, and also quite long.  My alternative proposal is to omit this section, and include the section I have on ''Beyond Hypothesis testing'' that follows after this}  
It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.  In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.  By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.

The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).  We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.  For this reason we generate the null distribution directly from the model as outlined above.   The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).  The plots in Fig. 3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.  Our approach replicate these findings for the $\tau$-statistic.  We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.

The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.  As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.  Fig. S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.  The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.  The reality is worse, we find it overlaps almost completely.  We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.

The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''

\subsection{Information criteria will not serve.}
One will commonly observe models representing alternative processes being compared in this way through 
the use of various information criteria such as the Akaike information criterion.  
While tempting to apply in this situation, such approaches are not suited to this problem for several reasons.  
The first is that information criteria are not concerned with the model choice objective we have in mind,
as they are typically applied to find an adequate model description without too many parameters that the system may be overfit.  
More pointedly, information criteria have no inherent notion of uncertainty.
Information criteria tests alone will not tell us our chances of a false alarm, of missing a real signal, 
or how much data we need to be confident in our ability to detect transitions.

\subsection{Beyond hypothesis testing}
\cb{If we remove the ``On a hypothesis-testing framework'' narrative above, I would then insert this.  I'm not sure if the wording of this next section is too casual.}
%Hypothesis testing only runs into trouble when we make the mistake of substituting uncertainty for certainty, that we imagine the phrase ``statistically significant'' is meaningful.  
The language of hypothesis testing is built around a strong bias that false positives are worse than false negatives, 
and consequently an a emphasis on $p$-values rather than power.  
In the context of early warning signals this is perilous -- 
it suggests that we would rather fail to predict a catastrophe than to sound a false alarm, 
while the question of when we have enough data to adequately detect an approaching transition
should be at least as important as the chance of a false alarm.

>>>>

%\documentclass[authoryear,5p]{elsarticle}
%\documentclass[authoryear,preprint,11pt]{elsarticle}
\documentclass[authoryear,review,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage[pdftex, colorlinks]{hyperref}
\usepackage{float}

\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in

\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

\newcommand{\cb}[1]{{\it (#1)}}

\journal{\tiny Journal of the Royal Society Interface} 
\begin{document}
\begin{frontmatter}
  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
  \author[cpb]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  \author[esp]{Alan Hastings}
  \cortext[cor1]{Corresponding author.}
  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}

@@

% rewrite

@@

% erg, a bit mor transition needed?

We discuss the limitations of current approaches that rely on summary statistics
and introduce a model based approach to detect early warning signals.
We illustrate how stochastic differential equation (SDE) models can be used 
to reflect the uncertainty inherent in the detection of early warning signals.
We caution against paradigms that are not useful for capturing this uncertainty in 
a model-selection based approach, such as information criteria,
and suggest reciever-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets 
and different early warning indicators have in detecting early warning signals.

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption 
that the system dynamics contain a saddle-node bifurcation by
looking for patterns that are associated with kind of transition.
Typical patterns include an increasing trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
While attractive for their simplicity, such approaches are subject to numerous challenges.
In this paper we argue for a model-based approach, 
and describe how this can be done in a way that best addresses these difficulties.
We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.

%%% hmm, what about all the red-herring topics, like change point analysis?  Add a discussion such as Contamin & Ellison ~\citep{Contamin2009} have about forecasting shifts, not post-hoc identification of shifts.

\subsubsection*{Arbitrary windows}
While superficially simple to use,
the calculation of these statistics is subject to several arbitrary decisions and assumptions.

A basic difficulty arises from the need to assume a time-series is ergodic 
while trying to test if it is not.  
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation? 
As true replicates are seldom avialable in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.  
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.  
\citet{Lenton2012} demonstrates that these differences can influence the results, 
and that the different choices each carry advantages and disadvantages.

In addition to introducing the challenge of selecting a window size, 
this ergodic assumption, 
that averaging over time is equivalent to averaging over replicate realizations,  
raises further difficulties.  
While appropriate for a system that is stationary, 
or changing slowly enough in the window that it may appear stationary,
the assumption is at odds with the original hypothesis 
that the system is approaching a saddle-node bifurcation.

Further, certain statistics such as the critical slowing down measured in autocorrelation 
require data that is evenly sampled in time.  
Interpolating from existing data to create evenly spaced points is particularly problematic,
as this introduces an artificial autocorrelation into the data.

\subsubsection*{No quantitative measures}
Summary statistics typically invoke qualitative patterns such as an increase in statistic $x$, 
rather than a quantitative measure of the early warning pattern.  
This makes it difficult to compare between signals or to
attribute a statistical significance to the detection.
Some authors have suggested Kendall's correlation coefficient, 
$\tau$, could be used to quantify an increase~\citep{Dakos2008, Dakos20011} 
in autocorrelation or variance.  
Other measures of increase, such as Pearson's correlation coefficient have also been proposed~\citep{Drake2010},
while most of the literature simply forgoes quantifying the increase or estimating significance.  
While adequate in experimental systems that can compare patterns between controls and replicates~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
any real-world application of these approaches must be useful on a single time-series of observations.
In these cases a quantitative definition of a statistically significant detection is essential. 
Using Kendall's $\tau$ we will illustrate how such a quantitative measure 
can be used to better capture the uncertainty in identifying the early warning pattern.

\subsubsection*{Problematic null models}
Specifying an appropriate null model is also difficult.  
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.  
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent,
but this is an assumption that is violated by the very experimental design:
temporal correlations will exist in any finely-enough samled time series, 
and moving windows introduce temporal correlations in the statistics.
Under such a test any adequately large data set will find a significant result,
regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
this destroys the natural autocorrelation in the time series.  
More promising parametric null models have been proposed, 
such as autoregressive models in~\citet{Dakos2008}, bringing us  
closer to a model-based approach with explicit assumptions.  
Others have looked for alternative summary statistics where 
reasonable null models are more readily available, 
such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedaticity.

%To characterize the power of an potential early warning signal we need more than the 
%distribution of some leading indicator under the null hypothesis of a stable system.  
%Very little work has addressed the distribution of any proposed indicator under the 
%actual test hypothesis, that is, direclty measured statistical power. 
%(~\citep{Contamin2009} use a measure of effect size, the log ratio of maximum value of differences between adjacent indicator values in the time-series, as a measure of power

%This is a particular concern given the tendency for warning signals to be found more often in large data sets, 
%which could be evidence that large data sets are required for successful detection~\citep{Inman2011, Scheffer2010,Held2004, Dakos2008, Scheffer2009, Guttal2008, Carpenter2011 Bestelmeyer2011},
%rathern than that large data sets have higher false-positive rates under certain metrics.
% Ellison Eco App paper -- calls it power but it is not
%
\subsubsection*{Hidden assumptions}
%Differing system dynamics
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in summary-statistic based approach.  
For instance, variance may increase for reasons that 
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.   
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by 
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

%Differing causes of the transition.
Even when we can exclude or ignore such dynamics and 
restrict ourselves to systems that can support a saddle-node bifurcation,
any early warning signals approach must further assume that a changing parameter 
has brought the system closer to the bifurcation.  This excludes at least three alternative
explanations for the transition.  
The first is that a large perturbation of the system state 
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.  
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast. 
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}  
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.  
The third scenario is that the system does pass through a saddle-node bifurcation, 
but rather than gradually and monotonically approaching the critical point, the 
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection*{Summary-statistic approaches have less statistical power.}
Methods for the detection of early warning signals are continually challenged by 
inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
The challenge of extrapolating into the future across a dramatic shift is immense;
there is essentially no historical president for successfully \& reliably predicting regime shifts
outside of controlled settings. % \cite Tetlock?
The signals one attempts to detect with these approaches are extremely weak.
Despite the widespread recognition of the this need for large data sets, 
there has been very few studies quantitative studies of power looking at how much data is actually required~\citep{Contamin2009}.
\cb{this is the only example that claims to measure statistical power.  I find it incredible that Ellison, contrary to his own ecological statistics textbook, decides to just invent an arbitrary statistic (log ratio of the maximum value in the statistic seen before and after onset of the perturbation) and call this a measure of ``statistical power.''}
The Neyman-Pearson Lemma guarantees us that the most powerful test between hypotheses compares the 
likelihood that the data was produced under each. This requires a model-based approach, which,
as we will see, can have substantially more power to detect early warning signals than summary-statistic based approaches.

@@

\subsection*{Early warning signals as model choice}
It may be useful to think of the detection of early warning signals as a problem of model choice, rather than one of pattern recognition.  
The model-choice based approach attempts to frame each of the possible scenarios as structurally different equations,
making the assumptions explicit.  
In any model choice problem, it is important to identify the goal of the exercise -- such as the ability to generalize, to immitate reality, or to predict~\citep{Levins1966}.  
In this case generality is more important than realism or predictive capability: 
we will write down a general model that is capable of approximating the wide class of models that experience a critical transition through a saddle node bifurcation,
and another general model that is capable of representing these same systems when they are not approaching a bifurcation.  
These may be thought of as the hypothesis and null hypothesis,
though they are in fact compound hypotheses: in each case, we must first estimate the model parameters from the data before we can compare the models.  
In this approach it is not assumed that ``reality'' is included in the models being tested per se, 
but that one of the models is a better approximation of the true dynamics than the other.  
System whose dynamics violate the assumptions common to both models, 
such as in the examples of ~\citet{Hastings2009} where systems exhibit sudden transitions without warning, 
fall outside the set of cases where this approach would be valid; 
though the inability of either model to match the system dynamics could be an indication of such a violation.

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter, which may be slowly varying in time.  
Transforming this canonical form to allow for an arbitrary mean $\theta$,
the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi(r_t)$.
We expand around the fixed point and express in terms of the stochastic dynamics:

@@

\subsection*{Likelihood calculations}
The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
\begin{equation}
\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
\end{equation}
For~\eqref{LSN} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
\begin{align}
 \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
  \label{general_moments}
\end{align}

@@

\subsection*{Comparing Models}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

Let $L_0$ be the likelihood function for model 0, 
let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) 
be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; 
and define $L_1$, $\theta_1$, $L_1$ similarly for model 1. 
The statistic we will use is $\delta$, 
defined to be twice the difference in log likelihood of observing the data under the two MLE models,
\begin{equation}\label{delta}
\delta = -2 (\log L_0 - \log L_1 ).
\end{equation}
This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.

%\section{On a hypothesis-testing framework}
%\cb{Here's a longer section on hypothesis testing that was originally in the appendix.  One of the Ecology Letters reviewers seemed very upset that this information was in the appendix and not in the main text, so here it is.  This section is quite specific in references to ~\citet{Dakos2008}, and also quite long.  My alternative proposal is to omit this section, and include the section I have on ''Beyond Hypothesis testing'' that follows after this}  
%It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.  In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.  By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.  
%
%The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).  We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.  For this reason we generate the null distribution directly from the model as outlined above.   The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).  The plots in Fig. 3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.  Our approach replicate these findings for the $\tau$-statistic.  We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.  
%
%The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.  As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.  Fig. S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.  The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.  The reality is worse, we find it overlaps almost completely.  We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test. 
%
%The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''  
%
%
%

\subsection*{Information criteria will not serve.}
One will commonly observe models representing alternative processes being compared in this way through 
the use of various information criteria such as the Akaike information criterion.  
While tempting to apply in this situation, such approaches are not suited to this problem for several reasons.  
The first is that information criteria are not concerned with the model choice objective we have in mind,
as they are typically applied to find an adequate model description without too many parameters that the system may be overfit.  
More pointedly, information criteria have no inherent notion of uncertainty.
Information criteria tests alone will not tell us our chances of a false alarm, of missing a real signal, 
or how much data we need to be confident in our ability to detect transitions.

\subsection*{Beyond hypothesis testing}
\cb{If we remove the ``On a hypothesis-testing framework'' narrative above, I would then insert this.  I'm not sure if the wording of this next section is too casual.}
%Hypothesis testing only runs into trouble when we make the mistake of substituting uncertainty for certainty, that we imagine the phrase ``statistically significant'' is meaningful.  
The language of hypothesis testing is built around a strong bias that false positives are worse than false negatives, 
and consequently an a emphasis on $p$-values rather than power.  
In the context of early warning signals this is perilous -- 
it suggests that we would rather fail to predict a catastrophe than to sound a false alarm, 
while the question of when we have enough data to adequately detect an approaching transition
should be at least as important as the chance of a false alarm.
